\documentclass{article}

\usepackage{amssymb}
\usepackage{verbatim}
\usepackage[francais,english]{babel}
\usepackage{graphicx}
\usepackage{graphics}
\usepackage{lineno}
\usepackage[latin1]{inputenc}
\usepackage[table]{xcolor}
\usepackage[]{natbib}
\usepackage[]{multirow}
\usepackage[]{a4wide}
\usepackage{fancyhdr}
\usepackage{lastpage}
\usepackage{amsmath}
\usepackage{lineno}
%\linenumbers

%\definecolor{light-gray}{gray}{0.8}
%\linenumbers
%\frontmatter
%\pagestyle{arabic}

\begin{document}
%\pagenumbering{arabic}
%\cfoot{\thepage/\pageref{LastPage}}
%\setcounter{page}{1}

\baselineskip 0.8cm

\title{{\tt TESS3}: Fast Inference of Spatial Population Structure and Genome Scans for Selection} 

\date{}

\author{Kevin Caye$^1$ \and Timo M. Deist$^1$ \and Helena Martins$^1$ \and Olivier Michel$^2$ \and Olivier Fran\c cois$^1$}

\maketitle

\begin{center}
$^1$Universit\'e Grenoble-Alpes, Centre National de la Recherche Scientifique, TIMC-IMAG UMR 5525, Grenoble, 38042, France.\\
$^2$Universit\'e Grenoble-Alpes, Centre National de la Recherche Scientifique, GIPSA-lab UMR 5216, Grenoble, 38042, France.\\
\end{center}

\noindent{Running Title: {\tt TESS3}: Inference of Spatial Population Structure\\}

\noindent{\bf Keywords:} Inference of Population Structure, Geographic Variation, 
Genome Scans for Selection, Control of False Discoveries.

\vspace{1cm}

{\noindent Corresponding Author: Olivier Fran\c cois \\
Université Grenoble-Alpes,\\ 
TIMC-IMAG,  UMR CNRS 5525,\\
Grenoble, 38042, France.\\
+334 56 52 00 25 (ph.)  \\
+334 56 52 00 55 (fax)  \\
 \texttt{olivier.francois@imag.fr}}


\clearpage
\newpage 
%
%
%
%
\begin{center}{\bf
Abstract}
\end{center}

Geography and landscape are important determinants of genetic variation in natural populations, and several ancestry estimation methods have been proposed to investigate population structure using genetic and geographic data simultaneously. Those approaches are often based on computer-intensive stochastic simulations, and do not scale with the dimensions of the data sets generated by high-throughput sequencing technologies. There is a growing demand for faster algorithms able to analyze genome-wide patterns of population genetic variation in their geographic context.
                                    

In this study, we present {\tt TESS3}, a major update of the spatial ancestry estimation program {\tt TESS}. By combining matrix factorization and spatial statistical methods, {\tt TESS3} provides estimates of ancestry coefficients with accuracy comparable to {\tt TESS} and with run-times much faster than the Bayesian version. In addition, the {\tt TESS3} program can be used to perform genome scans for selection, and separate adaptive from non-adaptive genetic variation using ancestral allele frequency differentiation tests. The main features of {\tt TESS3} are illustrated using simulated data and analyzing genomic data from European lines of the plant species {\it Arabidopsis thaliana}.

\clearpage
\newpage 


\input{introduction.tex}
\input{corps.tex}
\input{discussion.tex}



\begin{thebibliography}{31}
\providecommand{\natexlab}[1]{#1}
\input{babelbst.tex}
\newcommand{\Capitalize}[1]{\uppercase{#1}}
\newcommand{\capitalize}[1]{\expandafter\Capitalize#1}

\bibitem[{Alexander \& Lange(2011)}]{alexander2011linking}
Alexander DH, Lange K (2011) Enhancements to the {ADMIXTURE} algorithm for
  individual ancestry estimation.
\newblock \emph{BMC bioinformatics}, \textbf{12}, 246.

\bibitem[{Atwell \emph{\bbletal{}}(2010)Atwell, Huang, Vilhj{\'a}lmsson
  \emph{\bbletal{}}}]{atwell2010genome}
Atwell S, Huang YS, Vilhj{\'a}lmsson BJ, \emph{\bbletal{}} (2010) Genome-wide
  association study of 107 phenotypes in {\it \uppercase{a}rabidopsis thaliana}
  inbred lines.
\newblock \emph{Nature}, \textbf{465}, 627--631.

\bibitem[{Bazin \emph{\bbletal{}}(2010)Bazin, Dawson, \&
  Beaumont}]{bazin2010likelihood}
Bazin E, Dawson KJ, Beaumont MA (2010) Likelihood-free inference of population
  structure and local adaptation in a {B}ayesian hierarchical model.
\newblock \emph{Genetics}, \textbf{185}, 587--602.

\bibitem[{Belkin \& Niyogi(2003)}]{belkin2003laplacian}
Belkin M, Niyogi P (2003) Laplacian eigenmaps for dimensionality reduction and
  data representation.
\newblock \emph{Neural Computation}, \textbf{15}, 1373--1396.

\bibitem[{Benjamini \& Hochberg(1995)}]{benjamini1995controlling}
Benjamini Y, Hochberg Y (1995) Controlling the false discovery rate: a
  practical and powerful approach to multiple testing.
\newblock \emph{Journal of the Royal Statistical Society}, \textbf{57},
  289--300.

\bibitem[{Cai \emph{\bbletal{}}(2011)Cai, He, Han, \& Huang}]{cai2011graph}
Cai D, He X, Han J, Huang TS (2011) Graph regularized nonnegative matrix
  factorization for data representation.
\newblock \emph{IEEE Transactions on Pattern Analysis and Machine
  Intelligence}, \textbf{33}, 1548--1560.

\bibitem[{Cavalli-Sforza \emph{\bbletal{}}(1994)Cavalli-Sforza, Menozzi, \&
  Piazza}]{cavalli1994history}
Cavalli-Sforza LL, Menozzi P, Piazza A (1994) \emph{The History and Geography
  of Human Genes}.
\newblock Princeton University Press, USA.

\bibitem[{Chen \emph{\bbletal{}}(2007)Chen, Durand, Forbes, \&
  Fran{\c{c}}ois}]{chen2007bayesian}
Chen C, Durand E, Forbes F, Fran{\c{c}}ois O (2007) Bayesian clustering
  algorithms ascertaining spatial population structure: a new computer program
  and a comparison study.
\newblock \emph{Molecular Ecology Notes}, \textbf{7}, 747--756.

\bibitem[{Chung(1997)}]{chung1997spectral}
Chung FR (1997) \emph{Spectral Graph Theory}, \bblvol{} 92 of {\it Regional
  Conference Series in Mathematics}.
\newblock American Mathematical Society, USA.

\bibitem[{Devlin \& Roeder(1999)}]{devlin1999genomic}
Devlin B, Roeder K (1999) Genomic control for association studies.
\newblock \emph{Biometrics}, \textbf{55}, 997--1004.

\bibitem[{Durand \emph{\bbletal{}}(2009)Durand, Jay, Gaggiotti, \&
  Fran{\c{c}}ois}]{durand2009spatial}
Durand E, Jay F, Gaggiotti OE, Fran{\c{c}}ois O (2009) Spatial inference of
  admixture proportions and secondary contact zones.
\newblock \emph{Molecular Biology and Evolution}, \textbf{26}, 1963--1973.

\bibitem[{Epperson(2003)}]{epperson2003geographical}
Epperson BK (2003) \emph{Geographical Genetics (MPB-38)}.
\newblock Princeton University Press, USA.

\bibitem[{Fran{\c{c}}ois \emph{\bbletal{}}(2006)Fran{\c{c}}ois, Ancelet, \&
  Guillot}]{franccois2006bayesian}
Fran{\c{c}}ois O, Ancelet S, Guillot G (2006) Bayesian clustering using hidden
  {M}arkov random fields in spatial population genetics.
\newblock \emph{Genetics}, \textbf{174}, 805--816.

\bibitem[{Fran{\c{c}}ois \emph{\bbletal{}}(2008)Fran{\c{c}}ois, Blum,
  Jakobsson, \& Rosenberg}]{franccois2008demographic}
Fran{\c{c}}ois O, Blum MG, Jakobsson M, Rosenberg NA (2008) Demographic history
  of {E}uropean populations of {\it \uppercase{a}rabidopsis thaliana}.
\newblock \emph{PLoS Genetics}, \textbf{4}, e1000075.

\bibitem[{Fran{\c{c}}ois \& Durand(2010)}]{franccois2010spatially}
Fran{\c{c}}ois O, Durand E (2010) Spatially explicit {B}ayesian clustering
  models in population genetics.
\newblock \emph{Molecular Ecology Resources}, \textbf{10}, 773--784.

\bibitem[{Frichot \& Fran{\c{c}}ois(2015)}]{frichot2015lea}
Frichot E, Fran{\c{c}}ois O (2015) {LEA}: an {R} package for {L}andscape and
  {E}cological {A}ssociation studies.
\newblock \emph{Methods in Ecology and Evolution}, \textbf{6}, 925--929.

\bibitem[{Frichot \emph{\bbletal{}}(2014)Frichot, Mathieu, Trouillon, Bouchard,
  \& Fran{\c{c}}ois}]{frichot2014fast}
Frichot E, Mathieu F, Trouillon T, Bouchard G, Fran{\c{c}}ois O (2014) Fast and
  efficient estimation of individual ancestry coefficients.
\newblock \emph{Genetics}, \textbf{196}, 973--983.

\bibitem[{Holsinger \& Weir(2009)}]{holsinger2009genetics}
Holsinger KE, Weir BS (2009) Genetics in geographically structured populations:
  defining, estimating and interpreting $\uppercase{F}_{\rm st}$.
\newblock \emph{Nature Reviews Genetics}, \textbf{10}, 639--650.

\bibitem[{Hudson(2002)}]{hudson2002generating}
Hudson RR (2002) Generating samples under a {W}right--{F}isher neutral model of
  genetic variation.
\newblock \emph{Bioinformatics}, \textbf{18}, 337--338.

\bibitem[{Jay \emph{\bbletal{}}(2012)Jay, Manel, Alvarez
  \emph{\bbletal{}}}]{jay2012forecasting}
Jay F, Manel S, Alvarez N, \emph{\bbletal{}} (2012) Forecasting changes in
  population genetic structure of alpine plants in response to global warming.
\newblock \emph{Molecular Ecology}, \textbf{21}, 2354--2368.

\bibitem[{Kim \& Park(2011)}]{kim2011fast}
Kim J, Park H (2011) Fast nonnegative matrix factorization: an active-set-like
  method and comparisons.
\newblock \emph{SIAM Journal on Scientific Computing}, \textbf{33}, 3261--3281.

\bibitem[{Kimura \& Weiss(1964)}]{kimura1964stepping}
Kimura M, Weiss GH (1964) The stepping stone model of population structure and
  the decrease of genetic correlation with distance.
\newblock \emph{Genetics}, \textbf{49}, 561.

\bibitem[{Mal{\'e}cot(1948)}]{malecot1948mathematics}
Mal{\'e}cot G (1948) \emph{Les Math{\'e}matiques de l'H{\'e}r{\'e}dit{\'e}}.
\newblock Masson, Paris.

\bibitem[{Manel \emph{\bbletal{}}(2010)Manel, Joost, Epperson
  \emph{\bbletal{}}}]{manel2010perspectives}
Manel S, Joost S, Epperson BK, \emph{\bbletal{}} (2010) Perspectives on the use
  of landscape genetics to detect genetic adaptive variation in the field.
\newblock \emph{Molecular Ecology}, \textbf{19}, 3760--3772.

\bibitem[{Mochizuki \emph{\bbletal{}}(2005)Mochizuki, Harada, Inada
  \emph{\bbletal{}}}]{mochizuki2005arabidopsis}
Mochizuki S, Harada A, Inada S, \emph{\bbletal{}} (2005) The {\it
  \uppercase{a}rabidopsis} {WAVY} {GROWTH} 2 protein modulates root bending in
  response to environmental stimuli.
\newblock \emph{The Plant Cell}, \textbf{17}, 537--547.

\bibitem[{Pritchard \emph{\bbletal{}}(2000)Pritchard, Stephens, \&
  Donnelly}]{pritchard2000inference}
Pritchard JK, Stephens M, Donnelly P (2000) Inference of population structure
  using multilocus genotype data.
\newblock \emph{Genetics}, \textbf{155}, 945--959.

\bibitem[{Raj \emph{\bbletal{}}(2014)Raj, Stephens, \&
  Pritchard}]{raj2014faststructure}
Raj A, Stephens M, Pritchard JK (2014) fast{STRUCTURE}: variational inference
  of population structure in large {SNP} data sets.
\newblock \emph{Genetics}, \textbf{197}, 573--589.

\bibitem[{Segelbacher \emph{\bbletal{}}(2010)Segelbacher, Cushman, Epperson
  \emph{\bbletal{}}}]{segelbacher2010applications}
Segelbacher G, Cushman SA, Epperson BK, \emph{\bbletal{}} (2010) Applications
  of landscape genetics in conservation biology: concepts and challenges.
\newblock \emph{Conservation Genetics}, \textbf{11}, 375--385.

\bibitem[{Weir(1996)}]{weir1990genetic}
Weir (1996) \emph{Genetic {D}ata {A}nalysis {II}}.
\newblock Sinauer Associates Inc., Sunderland, MA.

\bibitem[{Wollstein \& Lao(2015)}]{wollstein2015detecting}
Wollstein A, Lao O (2015) Detecting individual ancestry in the human genome.
\newblock \emph{Investigative Genetics}, \textbf{6}, 1--12.

\bibitem[{Wright(1943)}]{wright1943isolation}
Wright S (1943) Isolation by distance.
\newblock \emph{Genetics}, \textbf{28}, 114.

\end{thebibliography}


\input{appendix.tex}

\clearpage
\newpage

\input{figuresAndTables.tex}

\end{document}

    
