\clearpage
\newpage


\section{Results}
\paragraph{Statistical errors} We used coalescent simulations of neutral
polymorphisms under spatial models of admixture to compare the statistical
errors of the AQP and APLS estimates with those of the {\tt tess3}
algorithm~\citep{Caye2016}. The ground truth for the $Q$-matrix (${\bf Q}_0$)
was computed from the mathematical model for admixture proportions used to
generate the data. For the $G$-matrix, the ground truth matrix (${\bf G}_0$) was
computed from the empirical genotype frequencies in the two population samples
before an admixture event. The root mean squared errors (RMSE) for the ${\bf Q}$
and ${\bf G}$ estimates decreased as the sample size and the number of loci
increased (Figure~\ref{fig:fig1}). For all algorithms, the statistical errors were generally
small when the number of loci was greater than $10$k SNPs. Those results
provided evidence that the three algorithms produced equivalent estimates of the
matrices ${\bf Q}_0$ and ${\bf G}_0$. The results also provided a check
that the APLS and {\tt tess3} algorithms converged to the same estimates as
those obtained after the application of the AQP algorithm, which is guaranteed
to converge mathematically.


\paragraph{The benefit of including spatial information in algorithms} Using
neutral coalescent simulations of spatial admixture, we compared the statistical
estimates obtained from a spatial algorithm (APLS) and a non-spatial algorithm
(sNMF, Frichot et al. 2014). For various levels of ancestral population
differentiation, estimates obtained from the spatial algorithm were more
accurate than for those obtained using non-spatial approaches (Figure~\ref{fig:fig2}). For
the larger samples, much finer population structure was detected with the
spatial method than with the non-spatial algorithm (Figure~\ref{fig:fig2}).

In simulations of outlier loci, we used the area under the precision-recall
curve (AUC) for quantifying the performances of tests based on the estimates of
ancestry matrices, {\bf Q} and {\bf G}. In addition, we computed AUCs for
$F_{\rm ST}$-based neutrality tests using truly ancestral genotypes. As they
represented the maximum reachable values, AUCs based on truly ancestral
genotypes were always higher than those obtained for tests based on
reconstructed matrices. For all values of the relative selection intensity, AUCs
were higher for spatial methods than for non-spatial methods (Figure~\ref{fig:fig3}, the
relative selection intensity is the ratio of migration rates at neutral and
adaptive loci). For high selection intensities, the performances of tests based
on estimates of ancestry matrices were close to the optimal values reached by
tests based on true ancestral frequencies. These results provided evidence that
including spatial information in ancestry estimation algorithms improves the
detection of signatures of hard selective sweeps having occurred in unknown
ancestral populations.

\paragraph{Sensitivity of estimates to spatial measurements.} Next, we used the
simulated data sets to evaluate the robustness of APLS estimates to inaccurate
measurements of spatial coordinates. To this aim, Gaussian noise was added to
truly observed geographic coordinates by considering values of the
noise-to-signal ratio ranging from 0 to 3. We computed variograms in all
cases, and found that the spatial signal was removed from simulations with a
noise-to-signal ratio of ??? ($\sigma \approx 0$), while the signal was still
observable with a noise-to-signal ratio of 1.
%% TODO paragraph quand fig OK
A noise-to-signal ratio of 1 increased statistical errors in the $Q$-matrix
estimates, but the RMSEs remained generally lower than for methods without
spatial coordinates ($\alpha = 0$ and sNMF, Figure~\ref{fig:fig2_5}). For a noise-to-signal ratio
of 3, errors were higher than for methods without spatial coordinates
(Figure~\ref{fig:fig2_5}). To a large extent, estimates from the APLS algorithm
were robust to uncertainty in spatial measurements. Standard graphical tests
such as a variogram analysis can help deciding whether our spatially explicit
algorithm is useful or not.


\paragraph{Runtime and convergence analyses} We subsampled a large SNP data set
for {\it A. thaliana} ecotypes to compare the convergence properties and
runtimes of the {\tt tess3}, AQP, and APLS algorithms. In those experiments, we
used $K = 6$ ancestral populations, and replicated 5 runs for each simulation.
For $n = 100-600$ individuals ($L = 50$k SNPs), the APLS algorithm required more
iterations (25 iterations) than the AQP algorithm (20 iterations) to converge to
its solution (Figure~\ref{fig:fig4}). This was less than for {\tt tess3} (30 iterations). For
$L = 10-200$k SNPs ($n = 150$ individuals), similar results were observed. For
$50$k SNPs, the runtimes were significantly lower for the APLS algorithm than
for the {\tt tess3} and AQP algorithms. For $L = 50$k SNPs and $n = 600$
individuals, it took on average 1.0 min for the APLS and 100 min for the AQP
algorithm to compute ancestry estimates. For {\tt tess3}, the runtime was on
average 66 min. For $L = 100$k SNPs and $n = 150$ individuals, it took on
average 0.6 min (9.0 min) for the APLS (AQP) algorithm to compute ancestry
estimates. For {\tt tess3}, the runtime was on average 1.3 min. For those
values of $n$ and $L$, the APLS algorithm implementation ran about 2 to 100
times faster than the other algorithm implementations.
 
\paragraph{Human data analysis} To evaluate a case of model misspecification
%% TODO
 
\paragraph{Application to European ecotypes of {\it Arabidopsis thaliana}} We
used the APLS algorithm to survey spatial population genetic structure and
perform a genome scan for adaptive alleles in European ecotypes of the plant
species {\it A. thaliana}. The cross validation criterion decreased rapidly from
$K=1$ to $K=3$ clusters, indicating that there were three main ancestral groups
in Europe, corresponding to geographic regions in Western Europe, Eastern and
Central Europe and Northern Scandinavia. For $K$ greater than four, the values
of the cross validation criterion decreased in a slower way, indicating that
subtle substructure resulting from complex historical isolation-by-distance
processes could also be detected (Figure~\ref{fig:fig5}). The spatial analysis provided an
approximate range of $\sigma = 150$km for the spatial variogram (Figure~\ref{fig:fig5}).
Figure~\ref{fig:map} displays the $Q$-matrix estimate interpolated on a geographic map of
Europe for $K = 6$ ancestral groups. The estimated admixture coefficients
provided clear evidence for the clustering of the ecotypes in spatially
homogeneous genetic groups.

\paragraph{Targets of selection in {\it A. thaliana} genomes} Tests based on the
$F^Q_{\rm ST}$ statistic were applied to the 241k SNP data set to reveal new
targets of natural selection in the {\it A. thaliana} genome. {\it A. thaliana}
occurs in a broad variety of habitats, and local adaptation to the environment
is acknowledged to be important in shaping its genetic diversity through
space~\citep{Hancock2011, Fournier-Level2011}. The APLS algorithm was run on the
1,095 European lines of {\it A. thaliana} with $K=6$ ancestral populations and
$\sigma = 1.5$ for the range parameter. Using the Benjamini-Hochberg algorithm to
control the FDR at the level $1\%$, the program produced a list of 12,701
candidate SNPs, including linked loci and representing 3\% of the total number
of loci. The top 100 candidates included SNPs in the flowering-related genes
SHORT VEGETATIVE PHASE (SVP), COP1-interacting protein 4.1 (CIP4.1) and FRIGIDA
(FRI) ($p$-values $< 10^{-300}$). These genes were detected by previous scans
for selection on this dataset~\citep{Horton2012}. We performed a gene ontology
enrichment analysis using AmiGO in order to evaluate which biological functions
might be involved in local adaptation in Europe. We found a significant
over-representation of genes involved in cellular processes (fold enrichment of
1.06, $p$-value equal to 0.0215 after Bonferonni correction).


